---
title: "Big Cojo Cam"
author: "Ryan Wang"
date: "2023-07-19"
output:
    html_document:
      toc: true
      toc_float: true
      code_folding: show
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(tidymodels)
library(dplyr)
library(here)
library(zoo)
library(tseries)
library(stats)
library(astsa)
library(forecast)
library(car)
library(MASS)
library(finalfit) 
library(ggplot2)
library(forecast)
library(lubridate)
library(astsa)
library(dynlm)
library(knitr)
library(nlme)

bc_hour_complete <- read.csv(here("data", "bc_hour_complete.csv"))
```

### Big Cojo Cam - Plotting and Analyzing the Time Series
```{r}
# take a subset for analysis  
# and split into training and testing set.

bc_hour_ts_total_1 <- ts(bc_hour_complete$number_of_objects[(19):(19+504+12)], frequency = 24)
bc_hour_ts_train_1 <- ts(bc_hour_complete$number_of_objects[(19):(19+504)], frequency = 24) 
bc_hour_ts_test_1 <- 
  ts(bc_hour_complete$number_of_objects[(524):(524+12)], frequency = 24, start = c(22, 1))


time_bc_hour_ts_train <- strptime(bc_hour_complete$time[(19):(504+19)], "%m/%d/%Y %H:%M")
tsplot(x=time_bc_hour_ts_train, y=bc_hour_ts_train_1, xlab = 'time', ylab = 'Count')
```

The variance of the time series seems to be constant, so the box cox transformation is not necessary.

```{r}
acf2(bc_hour_ts_train_1, max.lag = 100)
```

The ACF plot shows significant spikes at seasonal lags, which is a sign of seasonality. The seasonality can be removed by differencing. 

Differencing is a technique to remove the trend or seasonality from a time series data. It involves calculating the difference between consecutive observations to create a new series

First Difference (d = 1): To remove a trend, you take the difference between each observation and its previous observation. This is called first-order differencing:
$y_t' = y_t - y_{t-1}$

The new series, y_t', will have one less observation than the original series, as the first observation is dropped due to the differencing operation.

Second Difference (d = 2): If the first difference still contains a trend or is not yet stationary, you can take the difference between each observation of the first-differenced series and its previous observation. This is called second-order differencing:
$y_t'' = y_t' - y_{t-1}'$

```{r}
diff_bc_hour_ts_train <- diff(bc_hour_ts_train_1, 24)
tsplot(diff_bc_hour_ts_train, ylab = "Differenced Count")

adf.test(diff_bc_hour_ts_train)
Box.test(diff_bc_hour_ts_train, lag = 24, type = c("Box-Pierce"), fitdf = 0)
```
To further confirm that the time series is stationary, I perform the ADF test. To ensure that the stationary time series is not a white noise so that the modelling is meaningful, I perform the Box-Pierce test.

The p-value of both tests are less than 0.05, and thus we can reject the null hypothesis and take that the series is not a white noise and stationary.

### North Beach Fort Cam - Model identification
```{r}
acf2(diff_bc_hour_ts_train, main = "Big Cojo Cam" ) 
```
ACF shows a geometric decay over both seasonal and non-seasonal lags, which could be a sign of autoregression and seasonal autoregression.

PACF also shows a geometric decay over both seasonal and the non-seasonal lags, which could be a sign of Moving average and seasonal moving average.

Since it is a ARMA(SARMA) process , the exact values of p and q cannot be determined with the ACF/PACF plots.

### North Beach Fort Cam - Modelling

```{r}
# auto.arima(diff_bc_hour_ts_train, trace=TRUE) # Best model: ARIMA(1,0,0)(2,0,0)[24]
bc_hour_model_auto <-arima(diff_bc_hour_ts_train, order=c(0,0,0),seasonal = list(order=c(2,0,0),period=24))

#### Model diagnosis - auto.arima ####
# 1. Normality test
residual_bc_hour_model_auto <- residuals(bc_hour_model_auto)
hist(residual_bc_hour_model_auto,density=20,breaks=20,col="blue",xlab="",main = "Histogram of Residual",prob=TRUE)

# 2. Correlation test
acf2(residual_bc_hour_model_auto) 
# ACF plot / PACF plot show almost no significant spike.

Box.test(residual_bc_hour_model_auto, type = c("Box-Pierce"))
Box.test(residual_bc_hour_model_auto, lag = 24, type = c("Box-Pierce"))

# Thus, I conclude that the residual of this model is a Guassian white noise.
summary(bc_hour_model_auto)
```

I first try the model given by the auto.arima() function.

ACF and PACF plots show almost significant spikes at seasonal lags.
The Box-Pierce tests give a p-value of 0.01245 when the lag is 24. I reject the null and conclude that the residuals are not identically independently distributed; the residual of this model is a Guassian white noise.

Thus, I conclude that the model does not fully capture the seasonal component.

```{r}
bc_hour_model_1 <-arima(diff_bc_hour_ts_train, order=c(2,0,2),seasonal = list(order=c(2,0,3),period=24))

#### Model diagnosis - auto.arima ####
# 1. Normality test
residual_bc_hour_model_1 <- residuals(bc_hour_model_1)
hist(residual_bc_hour_model_1,density=20,breaks=20,col="blue",xlab="",main = "Histogram of Residual",prob=TRUE)

# 2. Correlation test
acf2(residual_bc_hour_model_1)
Box.test(residual_bc_hour_model_1, type = c("Box-Pierce"))
Box.test(residual_bc_hour_model_1, lag = 24, type = c("Box-Pierce"))
# Both the Box-Pierce test and the Ljung-Box test give p-vlaues greater than 0.05. I fail to reject the null 
 
# Thus, I conclude that the residual of this model is a Guassian white noise.
summary(bc_hour_model_1)
```

Then I try the model with a higher order of parameters.

ACF and PACF plots show almost no significant spike and the Box-Pierce tests give p-values greater than 0.05. I fail to reject the null and conclude that the residuals are identically independently distributed. 

Thus, the residual of this model is a Guassian white noise.

```{r}
bc_hour_model_2 <-arima(diff_bc_hour_ts_train, order=c(2,0,2),seasonal = list(order=c(1,0,1),period=24))

#### Model diagnosis - auto.arima ####
# 1. Normality test
residual_bc_hour_model_2 <- residuals(bc_hour_model_2)
hist(residual_bc_hour_model_2,density=20,breaks=20,col="blue",xlab="",main = "Histogram of Residual",prob=TRUE)

# 2. Correlation test
acf2(residual_bc_hour_model_2) 
Box.test(residual_bc_hour_model_2, type = c("Box-Pierce"))
Box.test(residual_bc_hour_model_2, lag = 24, type = c("Box-Pierce"))
# Both the Box-Pierce test and the Ljung-Box test give p-vlaues greater than 0.05. I fail to reject the null 
 
# Thus, I conclude that the residual of this model is a Guassian white noise.
summary(bc_hour_model_2)
```

Another model with different parameters.

ACF and PACF plots show almost no significant spike but the Box-Pierce test give a p-value of 0.0327 when the lag is 24.

Again, the model might not fully capture the seasonal component.

```{r}
### Summary of the created model.
result <- AIC(bc_hour_model_auto, bc_hour_model_1, bc_hour_model_2) 
models <- list(bc_hour_model_auto, bc_hour_model_1, bc_hour_model_2) 
result$BIC <- sapply(models, BIC) 
kable(result, digits = 2, align = "c")
```


```{r}
# Plot the fitted values with raw data
bc_hour_fitted_1 <- diff_bc_hour_ts_train - residual_bc_hour_model_1
ts.plot(diff_bc_hour_ts_train, main = 'model B, fitted vs actual')
points(bc_hour_fitted_1, type = "l", col = 2, lty = 2)

bc_hour_fitted_2 <- diff_bc_hour_ts_train - residual_bc_hour_model_2
ts.plot(diff_bc_hour_ts_train, main = 'model C, fitted vs actual')
points(bc_hour_fitted_2, type = "l", col = 2, lty = 2)
```

Here, I plot the fitted values of the two models with the transformed data.


### Big Cojo Cam - Forecast
```{r}
# Plot the forecast values with actual data
pred_1 <- sarima.for(bc_hour_ts_train_1, n.ahead=12, plot.all=F, p=2, d=0, q=2, P=2, D=1, Q=3, S=0)
lines(bc_hour_ts_test_1, col="blue")
points(bc_hour_ts_test_1, col="blue")
legend("bottomleft", pch=1, col=c("red", "blue"), legend=c("Predicted Values", "True Values"))

pred_2 <- sarima.for(bc_hour_ts_train_1, n.ahead=12, plot.all=F, p=2, d=0, q=2, P=1, D=1, Q=1, S=0)
lines(bc_hour_ts_test_1, col="blue")
points(bc_hour_ts_test_1, col="blue")
legend("bottomleft", pch=1, col=c("red", "blue"), legend=c("Predicted Values", "True Values"))
```




